---
title: "R Code And Tasks Chapter 5 (MAS 6003)"
author: "Witold Wolski"
date: "December 28, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# 6.1 Types of 2 way tables - response / control variables

## 6.1.1 Case(a): Skin cancer (melanoma) data - 2 response variables

__Cross sectional__ study of malignant melanoma.
 Both tumour type and site are __response variables__ because none
of the row or column totals were fixed in advance of the data collection.


```{r}
library(reshape2)
rm(list=ls())
load("data/MAS367-GLMs.RData", envir = e <- new.env())

Mela <- e$Mela
head(Mela)
dcast(Mela,tumour.type~ site, value.var="number")
```

## 6.1.2 Case(b) : Flu vaccine data - 1 response and 1 control variable

Patients were randomly assigned to the two groups (Placebo, Vaccine), and the response (levels of an antibody found in the blood six weeks after vaccination) was determined. Antibody level is the __response__ and vaccine group is a __controlled variable__ (with totals fixed by experimental design).

```{r}
vaccine <- e$vaccine
head(vaccine)
dcast(vaccine, treatment ~ response, value.var="count")
```

## 6.2.1 Association, Independence and Homogeneity


Indepence :

Case(a)

Probabilities of interest $\pi_{ij} = P(A=i, B=j)$

$$
P(A = i,B = j) = P(A = i) \times P(B = j)
$$


Case(b)

Probabilities of interest are conditional probablities $\pi{ij} = P(B=j|A=i)$.

The interest is in whether the probability distribution of the response (antibody level) is the same in each level of the controlled variable (drug group). If it doesn't depend on $i$ then we can write 

$$(\pi_{ij} = \pi_{.j})$$.

This is known as __homogeneity__.


# 6.3 Distribution for two-way tables

## 6.3.1 Case(a): two response variables

## 6.3.2 Case(b): one response variable

## 6.3.3 Case(c): independent poisson (no fixed margins).


## 6.3.4 Expected values

# 6.4 GLMs and two-way contingency tables

## 6.4.1 Natural hypothesis are log-linear models

## 6.4.2 Poisson log-linear modelling for two-way tables

## 6.4.3 Maximum likelihood estimation for $\pi_{ij}$ in case (a) and Task 19 and Task 20.

Task 19 Verify the maximum likelihood estimate for $\pi_{ij}$ for the A + B model for case(a).
Task 20 Verify the maximum likelihood estimate for $\pi_{ij}$ for the A + B model for case(b).


# 6.5 Interaction plots and examination of residues.

```interaction.plot``` honorable mentioned.

# 6.6 Analysis of the skin cancer data (case(a)) using log-linear models (Task 21)

## 6.6.1 Fitted values for the skin cancer data

* $A \times B$ saturated model.

```{r}
glm.sat <- glm(number ~ factor(tumour.type) * factor(site), family = poisson(log), data=Mela)
matrix(glm.sat$fitted.values,ncol=3)

```

* $A + B$ :independence

```{r}
glm.indep <- glm(number ~ factor(tumour.type) + factor(site), family = poisson(log), data=Mela)
matrix(glm.indep$fitted.values,ncol=3)

```

* $A$ : independence and the same probability for each column category

```{r}
glm.indepCol <- glm(number ~ factor(tumour.type), family = poisson(log), data=Mela)
matrix(glm.indepCol$fitted.values,ncol=3)

```

* $B$ : independence and the same probability for each row category

 Calculate the table of fitted values for the linear predictor containing B for case(a)

```{r}
glm.indepRow <- glm(number ~ factor(site), family = poisson(log), data=Mela)
res<-matrix(glm.indepRow$fitted.values,ncol=3)
rowSums(res)
colSums(res)
```

## 6.6.2

* $A\times B$ : saturated model

```{r}
void <- glm(number ~ factor(tumour.type) * factor(site), family = poisson(log), data=Mela)
```

## 6.6.3 Skin cancer data case(a) revisited

* 1  The test of independence based on the log-linear model $A + B$

```{r}
1 - pchisq(glm.indep$deviance,6)
qchisq(0.975,6)
```

* 2 The usual Pearson $\chi^2$ test for independence

```{r}
chisq.test(matrix(Mela$number,nrow=4))
1-pchisq(65.813,6)

```

* 5 

Deviance residuals see 3.6.4 in the notes.
Pearson residuals see Task 9 and page 91.

```{r}
plot(glm.indep$y,glm.indep$fitted.values)
sglmindep<-summary(glm.indep)
plot(sglmindep$deviance.resid , main="deviance residuals")
plot((glm.indep$y-glm.indep$fitted.values)/sqrt(glm.indep$fitted.values), main = "pearson residuals")

```

* interaction plot

```{r}
par(mfrow=c(1,2))
with(Mela, {
interaction.plot(site, tumour.type, log(number),las=2)
interaction.plot(tumour.type, site, log(number),las=2)

})

```

* removing first row or first column

remove cancer type A

```{r}
MelaNoA <- Mela[Mela$tumour.type!="A",]
glm.indepNoA <- glm(number ~ factor(tumour.type) + factor(site), family = poisson(log), data=MelaNoA)
glm.indepNoA
qchisq(0.975, 4)


```


remove site Head

```{r}
MelaNoHead <- Mela[Mela$site!="Head.and.Neck",]

glm.indepNoHead <- glm(number ~ factor(tumour.type) + factor(site), family = poisson(log), data=MelaNoHead)
glm.indepNoHead
qchisq(0.975, 3)
```

* 9 and 10 is about pooling the B, C, D or Trunk and Extremities

* 11

A log-linear model can be fitted which is additive in the factors, but includes a term for the (1,1) cell — an indicator variable for that cell (that is, treats it as an outlier).

```{r}
head(Mela)
MelaFix <- Mela
MelaFix$labelAHead <- rep(0,nrow(Mela))
MelaFix$labelAHead[1] <- 1
MelaFix$labelAHead <- as.factor(MelaFix$labelAHead)

glm.indepNoHead <- glm(number ~ factor(tumour.type) + factor(site) + factor(labelAHead), family = poisson(log), data=MelaFix)
glm.indepNoHead
```

### Task 22 Verify the analysis (See above)

### Task 23 


For the $4\times3$ table in Example 6.1.1, and the independence model, show directly (with-
out fitting a log-linear model) that $\mu_{11} = 5.780$, $e_{P,11} = 6.747$, and $e_{D,11} = 5.135$.

### Task 24
