---
title: "R Code And Tasks Chapter 7 (MAS 6003)"
author: "Witold Wolski"
date: "December 29, 2016"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Three way tables

* all three factors are responses

* one factor is fixed and 2 are responses

* two factors are fixed and one is an response

## 7.4.3 Example: ulcers and aspirin

```{r}
library(reshape2)
rm(list=ls())
load("data/MAS367-GLMs.RData", envir = e <- new.env())

aspirin <- e$aspirin

dcast(aspirin,type.of.ulcer + ulcer.or.not ~ aspirin.use, value.var = "count")
```

## 7.4.4


* $BC+A$ 

i - A (aspirin use)
j - B (ulcer)
k - C (case control)

$$
\pi_{ijk} = y_{i..}/n
$$
$$
\mu_{ijk} = n_{jk} y_{i..}/n
$$
then
$$
\mu_{212} = n_{12} \cdot y_{2..}/n = 68 \cdot 47/250
$$


```{r}
(BCA <- glm(count ~ aspirin.use + ulcer.or.not * type.of.ulcer,data=aspirin,family="poisson"))
matrix(BCA$fitted.values, ncol=2, byrow=T)
```

* $BC + AB$

$$
\pi_{ijk} = y_{ij.}/n_j
$$
$$
\mu_{ijk} = n_{jk} \cdot y_{ij.}/n_j
$$
then
$$
\mu_{212} = n_{.12} \cdot y_{21.}/n_{.1.} = 
68* (25+6)/(25+6+39+62) = 15.9697
$$


```{r}
(BCAB <- glm(count ~ aspirin.use*type.of.ulcer + ulcer.or.not * type.of.ulcer,data=aspirin,family="poisson"))
matrix(BCAB$fitted.values, ncol=2, byrow=T)
```


* $BC + AC$

$$
\pi_{ijk} = y_{i.k}/n_k
$$

$$
\mu_{ijk} =  n_{.jk} \cdot y_{i.k}/n_k
$$

then

$$
\mu_{212} = n_{.12} \cdot y_{2.2}/n_{..2} = 
68* (6+8)/(62 + 53 + 6 + 8) = 7.379845
$$

```{r}
(BCAC <- glm(count ~ aspirin.use*ulcer.or.not + ulcer.or.not * type.of.ulcer,data=aspirin,family="poisson"))
matrix(BCAC$fitted.values, ncol=2, byrow=T)
```


## 7.4.6 Model Selection


2.  The row (aspirin user) proportions of $25/ 64 = 0.39$ ($B_1, C_1$), $6/68 = 0.09$ ($B_1, C_2$ ), $8/57 = 0.14$ ($B_2, C_1$ ), $8/61 = 0.13$ ($B_2, C_2$ ), suggest that aspirin usage is different for the gastric ulcer cases ($B_1, C_1$) from the other three categories.


4. 

Compare with Deviance with

$$
\chi^2_{n-p}
$$
If model acceptable than deviance should be lower.

```{r}
qchisq(0.95,BCAC$df.residual)

```

but all the models have higher deviance.


### Task 27 
Verfiy the analysis in example 7.4.3 (See above)

### Task 28

By fitting the models in R, show that the binomial $(\eta \sim 1)$ analysis gives the same residual deviance (and hence model fit) as the log-linear $(\eta  \sim A + B * C)$ analysis for the ulcer/aspirin data.


```{r}
(BCA <- glm(count ~ aspirin.use + ulcer.or.not * type.of.ulcer,data=aspirin,family="poisson"))



BCSum <-rep(aggregate(count ~ ulcer.or.not + type.of.ulcer, aspirin, sum )$count,each=2)

props <- aspirin$count / BCSum
props1 <- props[aspirin$aspirin.use == "user"]
BCSum <-BCSum[aspirin$aspirin.use == "user"]
(Binom <- glm(props1 ~ 1,family="binomial", weights = BCSum))

```

The deviances are indeed equal.