---
title: "R Code And Tasks Chapter 5 (MAS 6003)"
author: "Witold Wolski"
date: "December 27, 2016"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Chapter 5 Poisson regression

# 5.1 Introduction

pdf of poisson

$$
\frac {\lambda ^{k}e^{-\lambda }}{k!}
$$


## 5.2.1 Example : AIDS deaths over time (Task 15)

### 1 plot :
```{r}
rm(list=ls())
load("data/MAS367-GLMs.RData", envir = e <- new.env())

AIDS <- e$AIDS
par(mfrow=c(1,2))
plot(AIDS$quarter, AIDS$deaths)
plot(AIDS$quarter, log(AIDS$deaths+1))

```

### 2 fit poisson with log link

```{r}
glm.lin <- glm(deaths ~ quarter, data=AIDS, family=poisson(link='log'))
tmp <-summary(glm.lin)
tmp$deviance
qchisq(0.95,glm.lin$df.residual)
qchisq(0.95,glm.lin$df.residual) < tmp$deviance

```

### 3 adding a quadratic term

```{r}
glm.quad <- glm(deaths ~ quarter + I(quarter^2), data=AIDS, family=poisson(link='log'))
summary(glm.quad)
qchisq(0.95,glm.lin$df.residual)

```

### 4 a line predictor on log(x)


```{r}
glm.logline <- glm(deaths ~ I(log(quarter)), data=AIDS, family=poisson(link='log'))
summary(glm.logline)
qchisq(0.95,glm.logline$df.residual)

plot(log(AIDS$quarter), log(AIDS$deaths+0.5))

```

### 5

Thus possible simple models are a line in logx or a quadratic in x, but there are reservations
about both.


# 5.3 Adjusting for exposure : offset (Task 16)

An explanation of offset which is brief and clear can be found here (but not in the lecture notes of MAS 6003).

[stack overflow](http://stats.stackexchange.com/questions/11182/when-to-use-an-offset-in-a-poisson-regression)

## 5.3.1 Example: Smoking and heart disease



### 1,2,3,4 death rates
```{r}
library(ggplot2)
library(gridExtra)

smoking <- e$smoking
smoking$rate <- smoking$deaths/smoking$person.years * 1e5
lapply(smoking,class)
smoking$smoke <- as.factor(smoking$smoke)
p1 <- ggplot(smoking, aes(age, rate, colour=smoke)) + geom_point()
p2 <- ggplot(smoking, aes((age), I(log(rate)), colour=smoke)) + geom_point()
p3 <- ggplot(smoking, aes(I(log(age)), I(log(rate)), colour=smoke)) + geom_point()
p4 <- ggplot(smoking, aes(age, deaths, colour=smoke)) + geom_point()


grid.arrange(p1,p2,p3,p4, ncol = 2)

```

### 5 The model

```{r}
mod.offset <- glm(deaths~ offset(log(person.years)) + smoke * age + I(age^2), family = poisson, data=smoking)
summary(mod.offset)

```

With smokers = 1 and 0 for nonsmokers:
for non-smokers:

$-19.7+0.36x^2-0.02x^2$

for smokers:

$-17.34+0.33x^2 -0.02x^2$


# 5.4 Non negative data with variance $\propto$ means (Task 17)

Compare the output from fitting a Poisson with log link and a line predictor on x to the
data in Example 5.2.1 with that obtained using the log link and assuming that the variance
is proportional to the mean.

```{r}
glm.lin <- glm(deaths ~ quarter, data=AIDS, family=poisson(link='log'))
summary(glm.lin)

```

```{r}
glm.quasi <- glm(deaths ~ quarter, data=AIDS, family=quasi(variance = "mu", link='log') )
summary(glm.quasi)

```

The only difference is in the "Std. Error" (see summary output above).


# 5.5 Further statistical modelling of count data

A train operator is reviewing the capacity of its trains between certain English cities. Their
particular interest is in modelling the number of train passengers starting their train journey
between 16:00 and 18:00 on weekdays between English cities.

* $i$ starting station
* $j$ destination
* $n_{ij}$ number of people living within 5 miles of either station i or j
* $r_{ij}$ number of passengers starting journey between 4 and 6 pm between station i and j

$$
Po(\mu_{ij}) = \frac{\mu_{ij}^{r_{ij}}\exp(-\mu_{ij})}{r_{ij}!}
$$

$$
\mu_{ij} = \alpha_i \exp(\beta n_{ij})
$$


$$
L(\mu_{ij}) = \prod_i \prod_j \frac{\mu_{ij}^{r_{ij}}\exp(-\mu_{ij})}{r_{ij}!}
$$

$$
\begin{aligned}
l(\mu_{ij}) &= \sum_i \sum_j r_{ij} \log(\mu_{ij}) - (\mu_{ij}) - constant\\
&= \sum_i \sum_j r_{ij} (\log(\alpha_i) + \beta n_{ij}) - \alpha_i \exp(\beta n_{ij}) - constant
\end{aligned}
$$

Models the probablity of having $r_{ij}$ passengers starting journey from station $i$ to station $j$.

* S - starting station (same as i)
* P - number of people living within 5 miles of the starting or destination stations
* C - number of passengers starting their train journey between 4 and 6 pm at station S.



```{r}
train <- e$train
train.glm <- glm(C ~ factor(S) + P, family=poisson("log"),data=train)
summary(train.glm)
```

We want to estimate $\alpha_i$ and $\beta$ from the data to estimate $\mu_{ij}$ given $n_{ij}$.

* $\hat{\alpha_1} = \exp{\gamma_1}$
* $\hat{\alpha_i} = \exp{\gamma_1 + \gamma_i}$, for $i\ge 1$
* $\beta = \delta$

## Task 18

Use the output to determine $\alpha_1$. Hence, show that the parameter estimates $\alpha_i$ and $\beta$ satisfy the first of the mle equations
$$
\frac{\partial l(\mu_{ij})}{\partial \alpha_i} =  \sum_{j\in D(i)} [r_{ij}/\alpha_i- \exp(\beta n_{ij})] = 0
$$

```{r}
coefTrain <- coefficients(train.glm)
coefTrain
```

```{r}
alpha_1 = exp(coefTrain[1])
alpha_2_3 = exp(coefTrain[1]+coefTrain[2:3])
alphas <- c(alpha_1,alpha_2_3)
alphas
beta = coefTrain[4]
beta

```


```{r}
r1 <- subset(train, S == 1, select=c("C"))
n1 <- subset(train, S == 1, select = c("P"))
sum(r1 / alphas[1] - exp(beta * n1))


r2 <- subset(train, S == 2, select=c("C"))
n2 <- subset(train, S == 2, select = c("P"))
sum(r2 / alphas[2] - exp(beta * n2))


r3 <- subset(train, S == 3, select=c("C"))
n3 <- subset(train, S == 3, select = c("P"))
sum(r3 / alphas[3] - exp(beta * n3))

```

$$
\begin{aligned}
\frac{\partial l(\mu_{ij})}{\partial \beta} &=  (\sum_i \sum_{j \in D(i)} r_{ij} (\log(\alpha_i) + \beta n_{ij}) - \alpha_i \exp(\beta n_{ij}) - constant)'\\
&=  (\sum_i \sum_{j \in D(i)} r_{ij} \log(\alpha_i) + r_{ij}\beta n_{ij} - \alpha_i \exp(\beta n_{ij}) - constant)'\\
&=  \sum_i \sum_{j \in D(i)} r_{ij}n_{ij} - n_{ij}\alpha_i \exp(\beta n_{ij})\\
&= \sum_i \sum_{j \in D(i)} n_{ij}(r_{ij} - \alpha_i \exp(\beta n_{ij})) = 0
\end{aligned}
$$
It can be seen that this sums to 0 since the term $r_{ij} - \alpha_i \exp(\beta n_{ij})$ sums up to zero as we have shown above.


